A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-07-15 Manuscript/PNAS-Adams-rticle.Rmd
pnas vsn

author:
  - name: Dean C. Adams
    affiliation: a
  - name: Erica K. Baken
    affiliation: a,b
  - name: Michael L. Collyer
    affiliation: b
address:
  - code: a
    address: Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, Iowa, USA.
  - code: b
    address: Department of Science, Chatham University, Pittsburgh, Pennsylvania, USA.

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Adams et. al

# Place DOI URL or CRAN Package URL here
doi_footer: "doi: TBD"

# Abstract
abstract: |
  Macroevolutionary studies frequently characterize the phylogenetic signal in phenotypes, and wish to compare the strength of that signal across traits. However, analytical tools for such comparisons have largely remained underdeveloped. Here we evaluate the efficacy of one commonly used parameter (Pagel's $\lambda$) to estimate the strength of phylogenetic signal in phenotypic traits, and evaluate the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal. We find that $\lambda$ behaves as a Bernoulli random variable, and that estimates are increasingly skewed at larger and smaller input levels of phylogenetic signal. Further, the precision of $\lambda$ in estimating actual levels of phylogenetic signal is often inaccurate, and biological interpretations of the strength of phylogenetic signal based on $\lambda$ are therefore compromised. We propose a standardized effect size based on $\kappa$, ($Z_\kappa$), which measures the strength of phylogenetic signal more reliably than does $\lambda$, and places that signal on a common scale for statistical comparison. We develop tests based on $Z_\kappa$ to provide a mechanism for formally comparing the strength of phylogenetic signal across datasets, in much the same manner as effect sizes may be used to summarize patterns in quantitative meta-analysis. Our approach extends the phylogenetic comparative toolkit to address hypotheses that compare the strength of phylogenetic signal between various phenotypic traits, even when those traits are found in different evolutionary lineages or have different units or scales.

# Significance
significance: |
  Evolutionary biologists wish to quantify and compare the strength of phylogenetic signal across traits, but analytical tools for these comparisons are generally lacking. Here we develop a standardized effect size based on $\kappa$ ($Z_\kappa$), which measures the strength of phylogenetic signal on a common statistical scale, and provides a mechanism for formally comparing the strength of phylogenetic signal across datasets. Additionally, we find that a commonly used parameter (Pagel's $\lambda$) is insuitable for this purpose. Our procedure enables biologists to quantitatively address hypotheses that compare the strength of phylogenetic signal between various phenotypic traits, even when those traits are found in different evolutionary lineages or have different units or scales.

# Optional: Acknowledgements
acknowledgements: |
   This work was supported in part by NSF grant DBI-1902511 (to D.C.A.) and DBI-1902694 (to M.L.C.).

# Optional: One or more keywords
keywords: phylogenetic signal, macroevolution, lambda, kappa

# Paper size for the document, values of letter and a4
papersize: letter

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 9pt

# Optional: Force one-column layout, default is two-column
#one_column: true

# Optional: Enables lineo mode, but only if one_column mode is also true
lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
# numbersections: true

# Optional: Specify the depth of section number, default is 5
secnumdepth: 3

# Optional: Bibliography 
csl: pnas.csl
bibliography: Lambda-Refs.bib

# Optional: include-after
#include-after: somefile.tex

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Enable a 'Draft' watermark on the document
watermark: true

# Produce a pinp document
# 
# Other options that can be set here, shown with their defaults
#   keep_tex: TRUE
#   collapse: FALSE
#   citation_package: natbib
#
# Any other arguments to rmarkdown::pdf_document could be set 
# here as well
#
output:
  pinp::pinp:
#    includes:
#      after_body: bib.tex

---

# Introduction

Investigating macroevolutionary patterns of trait variation requires a phylogenetic perspective, because the shared ancestry among species violates the assumption of independence among trait values that is common for statistical tests [@Felsenstein1985; @HarveyPagel1991]. Accounting for this evolutionary non-independence is the purview of *phylogenetic comparative methods* (PCMs): a suite of analytical tools that condition trends in the data on the phylogenetic relatedness of observations [@Grafen1989; @GarlandIves2000; @Rohlf2001; @ButlerKing2004; @MartinsHansen1997; @OMeara_et_al2006; @RevellHarmon2008; @Beaulieu_et_al2012; @Adams2014b; @AdamsCollyer2018b]. These methods are predicated on the notion that phylogenetic signal -- the tendancy for closely related species to display similar trait values -- is present in cross-species datasets [@Felsenstein1985; @Pagel1999; @Blomberg_et_al2003]. Indeed, under numerous evolutionary models, phylogenetic signal is to be expected, as stochastic character change along the hierarchical structure of the tree of life generates trait covaration among related taxa [@Felsenstein1985; @Blomberg_et_al2003; @Revell_et_al2008].

@@

address hypotheses that compare the strength of phylogenetic signal between various phenotypic traits, even when those traits are found in different evolutionary lineages or have different units or scales

@@

# Code

Pandoc segments are highlighted as usual, and per a convention in this
template also shown in a framed and slightly shaded box as seen here and
above.

Another example from Python:

```py
# A Python example
def printSomething(str):
   "This prints the string passed in"
   print str

```

<!-- Acknowledgements, and References, are automatically typeset after this -->

>>>>

# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
author:
  - name: Dean C. Adams
    affiliation: a
  - name: Erica K. Baken
    affiliation: a,b
  - name: Michael L. Collyer
    affiliation: b
address:
  - code: a
    address: Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, Iowa, USA.
  - code: b
    address: Department of Science, Chatham University, Pittsburgh, Pennsylvania, USA.

corresponding_author:
  - code: 2
    text: "To whom correspondence should be addressed. E-mail: dcadams@iastate.edu"

# For footer text
lead_author_surname: Adams et al.

author_contributions: |
  D.C.A. designed the research; D.C.A., E.K.B., and M.L.C. performed the research and wrote the paper.

## Remove this if not required
conflict_of_interest: |
  The authors declare no conflict of interest.

abstract: |
  Macroevolutionary studies frequently characterize the phylogenetic signal in phenotypes, and wish to compare the strength of that signal across traits. However, analytical tools for such comparisons have largely remained underdeveloped. Here we evaluate the efficacy of one commonly used parameter (Pagel's $\lambda$) to estimate the strength of phylogenetic signal in phenotypic traits, and evaluate the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal. We find that $\lambda$ behaves as a Bernoulli random variable, and that estimates are increasingly skewed at larger and smaller input levels of phylogenetic signal. Further, the precision of $\lambda$ in estimating actual levels of phylogenetic signal is often inaccurate, and biological interpretations of the strength of phylogenetic signal based on $\lambda$ are therefore compromised. We propose a standardized effect size based on $\kappa$, ($Z_\kappa$), which measures the strength of phylogenetic signal more reliably than does $\lambda$, and places that signal on a common scale for statistical comparison. We develop tests based on $Z_\kappa$ to provide a mechanism for formally comparing the strength of phylogenetic signal across datasets, in much the same manner as effect sizes may be used to summarize patterns in quantitative meta-analysis. Our approach extends the phylogenetic comparative toolkit to address hypotheses that compare the strength of phylogenetic signal between various phenotypic traits, even when those traits are found in different evolutionary lineages or have different units or scales.

significance: |
  Evolutionary biologists wish to quantify and compare the strength of phylogenetic signal across traits, but analytical tools for these comparisons are generally lacking. Here we develop a standardized effect size based on $\kappa$ ($Z_\kappa$), which measures the strength of phylogenetic signal on a common statistical scale, and provides a mechanism for formally comparing the strength of phylogenetic signal across datasets. Additionally, we find that a commonly used parameter (Pagel's $\lambda$) is insuitable for this purpose. Our procedure enables biologists to quantitatively address hypotheses that compare the strength of phylogenetic signal between various phenotypic traits, even when those traits are found in different evolutionary lineages or have different units or scales.

acknowledgements: |
  We thank E. Glynne and B. Juarez for comments on early drafts of the manuscript. This work was supported in part by NSF grant DBI-1902511 (to D.C.A.) and DBI-1902694 (to M.L.C.). \hfill\break

keywords:
  - phylogenetic signal
  - macroevolution
  - lambda
  - kappa

## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
pnas_type: pnasresearcharticle

bibliography: Lambda-Refs.bib
csl: pnas.csl

## change to true to add optional line numbering
lineno: true

output: rticles::pnas_article
---
Investigating macroevolutionary patterns of trait variation requires a phylogenetic perspective, because the shared ancestry among species violates the assumption of independence among trait values that is common for statistical tests [@Felsenstein1985; @HarveyPagel1991]. Accounting for this evolutionary non-independence is the purview of *phylogenetic comparative methods* (PCMs): a suite of analytical tools that condition trends in the data on the phylogenetic relatedness of observations [@Grafen1989; @GarlandIves2000; @Rohlf2001; @ButlerKing2004; @MartinsHansen1997; @OMeara_et_al2006; @RevellHarmon2008; @Beaulieu_et_al2012; @Adams2014b; @AdamsCollyer2018b]. These methods are predicated on the notion that phylogenetic signal -- the tendancy for closely related species to display similar trait values -- is present in cross-species datasets [@Felsenstein1985; @Pagel1999; @Blomberg_et_al2003]. Indeed, under numerous evolutionary models, phylogenetic signal is to be expected, as stochastic character change along the hierarchical structure of the tree of life generates trait covaration among related taxa [@Felsenstein1985; @Blomberg_et_al2003; @Revell_et_al2008].

@@

**To be edited.** address hypotheses that compare the strength of phylogenetic signal between various phenotypic traits, even when those traits are found in different evolutionary lineages or have different units or scales.

It is common in comparative evolutionary studies to characterize the phylogenetic signal in phenotypic traits to determine the extent to which shared evolutionary history has generated trait covariation among taxa. However, while numerous analytical approaches may be used to quantify phylogenetic signal [e.g., @Abouheif1999; @Gittleman1990; @Blomberg_et_al2003; @Adams2014a; @Pagel1999], methods that explicitly measure the strength of phylogenetic signal, or facilitate comparisons among datasets, have remained underdeveloped. In this study, we evaluated the precision of one common measure, Pagel's $\lambda$, and explored its efficacy for characterizing the strength of phylogenetic signal in phenotypic data. Using computer simulations, we found that the precision of $\lambda$ increased with increasing sample sizes; a pattern noted previously [@Boettiger_et_al2012], and one that conformed with parametric statistical theory [@Cohen1988]. However, we also found that vastly different $\lambda$ estimates could be obtained from data containing the same level of phylogenetic signal, and that similar $\lambda$ estimates may be obtained from data containing differing levels of phylogenetic signal. Further, the precision of $\lambda$ varied with the strength of phylogenetic signal, where lower precision was observed when in data whose phylogenetic signal was of intemediate strength. From these findings we conclude that $\lambda$ is not a reliable indicator of the observed strength of phylogenetic signal in phenoytpic datasets, and that biological interpretations of the strength of signal based on this parameter may innacurately characterize such effects.

As an alternative, we described a standardized effect size ($Z$) for assessing the strength of phylogenetic signal. $Z$ expresses the magnitude of phylogenetic signal as a standard normal deviate, which is easily interpretable as the strength of phylogenetic signal relative to the mean. We applied this concept to both $\lambda$ and $\kappa$, and found that $Z_\kappa$ was a better estimate of the strength of phylogenetic signal in phenotypic data. First, $Z_\kappa$ was more precise than $Z_\lambda$, and precision was more consistent across the range of input levels of phylogenetic signal. Additionally, values of $Z_\kappa$ more accurately tracked known changes in the magnitude of phylogenetic signal, as demonstrated by the linear relationship between $Z_\kappa$ and $\lambda_{in}$. Thus, $Z_\kappa$ holds promise as a measure of the relative strength of phylogenetic signal that reflects the magnitude of this effect in phenotypic data. We therefore recommend that future studies interested in the strength of phylogenetic signal incorporate $Z_\kappa$ as a statistical measure of this effect.

Based on the effect size $Z_\kappa$, we then proposed a two-sample test, which provides means of determining whether the strength of phylogenetic signal is greater in one phenotypic trait as compared to another, via a hypothesis test. Prior studies have summarized patterns of variation in phylogenetic signal across datasets using summary test values, such as $\kappa$ [e.g.,  @Blomberg_et_al2003]. However, $\kappa$ does not scale linearly with input levels of phylogenetic signal, and its variance increases (i.e., precision decreases) with increasing strength of phylogenetic signal [@Munkemuller_et_al2012; @DinizFilho2012: see also Supporting Information]. Thus, $\kappa$ should not be considered an effect size that measures the strength of phylogenetic signal on a common scale. By contrast, standardizing $\kappa$ ($Z_\kappa$, via equation 2) alleviates these concerns, and facilitates formal statistical comparisons of the strength of signal across datasets. Thus when viewed from this perspective, the approach developed here aligns well with other statistical approaches such as meta-analysis [sensu @HedgesOlkin1985; @Glass1976; @Arnqvist1995], where summary statistics across datasets are converted to standardized effect sizes for subsequent "higher order" statistical summaries or comparisons. As such, our approach enables evolutionary biologists to quantitatively examine the relative strength of phylogenetic signal across a wide range of phenotypic traits, and thus opens the door for future discoveries that inform on how phenotypic diversity accumulates in macroevolutionary time across the tree of life.

One important advantage of the approach advocated here is that the resulting effect sizes ($Z_\kappa$) are dimensionless, as the units of measurement cancel out during the calculation of $Z$ [@SokalRohlf2012]. Thus, $Z_\kappa$ represents the strength of phylogenetic signal on a common and comparable scale -- measured in standard deviations -- regardless of the initial units and original scale of the phenotypic variables under investigation. This means that the strength of phylogenetic signal may be compared across datsets for continuous phenotypic traits measured in different units and scale, because those units have been standardized through their conversion to $Z_\kappa$. For example, our approach could be utilized to determine whether the strength of phylogenetic signal (say, in response to ecological differentiation) is stronger in morphological traits (linear traits: $mm$), physiological traits (metabolic rate: $\frac{O^2}{min}$), or behavioral traits (aggression: $\frac{\#{displays}}{second}$). In fact, our empirical example provided such a comparison, as SA:V is represented in $mm^{-1}$ while relative body size is a unitless ratio ($\frac{BW}{SVL}$). Additionally, our method  is capable of comparing the strength of phylogenetic signal in traits of different dimensionality, as estimates of phylogenetic signal using $\kappa$ have been generalized for multivariate data [$\kappa_{mult}$: see @Adams2014a]. Furthermore, tests based on $\hat{Z}_{12}$ may be utilized for comparing the strength of phylogenetic signal among datasets containing a different number of species, and even for phenotypes obtained from species in different lineages, because their phylogenetic non-independence and observed variation are taken into account in the generation of the empirical sampling distribution via permutation.

This study is not the first to compare $\lambda$ and $\kappa$ for their ability as statistics to measure phylogenetic signal.  Our results for $\lambda$ and $\kappa$ values are consistent with those found in the simulations performed by Münkemüller et al. [-@Munkemuller_et_al2012], but that study investigated type I error rates and statistical power, finding that $\lambda$ performed better in both regards, irrespective of species number in trees.  Although not the central focus of their study, the same tendency for variable $\lambda$ and consistent $\kappa$ at intermediate phylogenetic signal strengths was observed [see Fig. 2, @Munkemuller_et_al2012].  Recent work by Molina-Venegas and Rodríguez [-@MolinaVenegas2017] found that $\kappa$ but not $\lambda$ tended to inflate the estimate of phylogenetic signal, leading to moderate type I and type II biases, if polytomic chronograms were used.  Their work more thoroughly addressed previous observations of inflated $\kappa$ for incompletely resolved phylogenetic trees [@Davies2012; @Munkemuller_et_al2012].  An interesting question is whether an inflated $\kappa$ value leads to an inflated $Z_\kappa$ or does a tendency of a particular tree to inflate estimates of $\kappa$ also inflate the values in random permutations of a test, in which case $Z_\kappa$ is robust to polytomies?  We repeated the analyses in Figure 4, adjusting trees to have 50% collapsed nodes, per the technique of Molina-Venegas and Rodríguez [-@MolinaVenegas2017], and found results were consistent (see Supporting Information).  This confirms that any tendency of incompletely resolved trees to inflate $\kappa$ as a descriptive statistic does not inflate $Z_\kappa$ as an effect size. Furthermore, because comparison of effect sizes in a test is a comparison of locations of observed values in their sampling distributions, which would shift concomitantly because of this tendency, the $Z_{12}$ test statistic in equation 4 appears to be robust in spite of unresolved trees.

Phylogenetic signal can be thought of as both an attribute to be measured in the data and a parameter that can be tuned to account for the phylogenetic non-independence among observations, for analysis of the data.  As such, $\lambda$ is appealing, as a statistic that potentially fulfills both roles.  However, the inability to estimate phylogenetic signal with $\lambda$ for data simulated with known phylogenetic signal is troublesome, and we recommend evolutionary biologists refrain from viewing it as a statistic to describe the amount of phylogenetic signal in the data.  Interestingly, $\kappa$ -- when standardized to an effect size $Z_\kappa$ -- is a better statistic for measuring the amount of phylogenetic signal in data simulated with respect to known levels of $\lambda$.  Although $\lambda$ might be viewed as an important parameter for modifying the the conditional estimation of linear model coefficients with respect to phylogeny, it is neither a statistic that has meaningful comparative value as a measure of phylogenetic signal nor a statistic that lends itself well to reliable calculation of a test statistic.  By contrast, $\kappa$ has been shown here to be a reliable statistic, but only when standardized by the mean and standard deviation of its empirical sampling distribution (i.e., when converted to the effect size, $Z_\kappa$).  Because one has control over the number of permutations used in analysis, one can be assured with many permutations that the empirical sampling distribution is representative of true probability distributions [@AdamsCollyer2018b].  With low coefficients of variation for $Z_\kappa$ (Figure 4), it is difficult to imagine that a hypothesis test can improve equation 4 for efficiently comparing phylogenetic signal for different traits, different trees, or a combination of both.

@@

<!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
\showmatmethods
\showacknow
\pnasbreak
